---
title: "Plan annotation effort, costs, and pilot precision"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Plan annotation effort, costs, and pilot precision}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 95)
```

This tutorial illustrates the study-planning additions in development source
0.4.1.9000. It does not change the qualified 0.4.1 archive or establish a new
release. All observations and monetary amounts below are synthetic. No requests
are executed and no provider prices are retrieved.

| Quantity | Purpose |
|---|---|
| Distinct pilot items | Information for estimating source variation and coefficients |
| Annotations per item | The averaging protocol whose reliability is projected |
| Items per request | Shared context and request overhead; not additional repeated measurements |

## Fit a small Gaussian pilot

The first example uses independent-item annotations with item, rater and
item-by-rater sources. Runs supply replication of those sources through distinct
coded cells. This working model defines the example's intended score universe.

```{r pilot}
library(Gtheory4LLM)
set.seed(2718)
panel <- expand.grid(item = 1:24, rater = 1:4, run = 1:3,
                     KEEP.OUT.ATTRS = FALSE)
item_rater <- as.integer(interaction(panel$item, panel$rater, drop = TRUE))
panel$quality <- rnorm(24)[panel$item] + rnorm(4, sd = 0.5)[panel$rater] +
  rnorm(96, sd = 0.4)[item_rater] + rnorm(nrow(panel), sd = 0.8)
design <- gt_design("item", c("rater", "run"),
                    random = ~ item + rater + item:rater)
control <- gt_control(gaussian = list(extra_tries = 2L, retry_seed = 439L))
fit <- gt_fit(panel, "quality", design, control = control)
stopifnot(isTRUE(fit$numerically_accepted))
gt_reliability(fit)$per_trait
```

Acceptance permits the supported projections. It does not establish a global
optimum, adequate pilot precision, labeling accuracy or correct source selection
for another study.

## Compare costs within a declared allocation grid

Here one annotation costs 0.01 illustrative dollars, with a setup cost of two
dollars per rater. The cost callback could instead return separate request,
token, human-review and expected-retry costs, provided all assumptions use the
same currency and are recorded. Costs do not introduce a statistical benefit
from review or change the population of raters represented by the fit.

```{r monetary-plan}
plan <- gt_plan(fit,
  candidates = list(rater = c(2, 4, 6), run = c(1, 3, 5)),
  cost = function(d) list(
    costs = data.frame(annotation = 0.01 * d$total_annotations,
                       setup = 2 * d$allocation_rater),
    resources = data.frame(requests = d$total_annotations),
    assumptions = "Illustrative prices; one item per request."),
  currency = "USD", cost_date = "2026-10-09", study_items = 1000,
  target = 0.80, coefficient = "Phi", baseline = 1L,
  constraints = function(d) data.frame(budget = d$total_cost <= 180))
as.data.frame(plan)[, c("design_id", "total_cost", "screening_value",
                       "feasible", "minimum_cost", "pareto")]
head(plan$source_contributions)
```

Minimum cost means minimum among the supplied candidates that meet both the
constraints and the selected reliability screen. All candidates remain in the
result, including cost ties and infeasible rows. Pareto candidates have no
feasible cheaper-or-equal alternative with equal-or-better performance and at
least one strict improvement. `screening = "lower"` separately uses an available
lower confidence bound. It never replaces a missing bound with a point estimate;
these are pointwise intervals, not post-search or future-panel guarantees.

`study_items` changes total cost, not the per-item coefficient. Every candidate
retains the same declared random/fixed universe. In particular, the planner
cannot make a protocol appear more dependable by changing a random facet into
a fixed one.

## Make the batch target explicit

The next synthetic panel adds a shared shift for each call of six items. Every
condition uses the same grouping. A recorded call identifier allows the fit to
retain that layout even when rows are reordered.

```{r batch-fit}
batched <- panel
group <- (batched$item - 1L) %/% 6L + 1L
call_index <- as.integer(interaction(group, batched$rater, batched$run, drop = TRUE))
batched$call_id <- paste0("call_", call_index)
batched$quality <- batched$quality + rnorm(max(call_index), sd = 0.7)[call_index]
batch_design <- gt_design("item", c("rater", "run"),
  random = ~ item + rater + item:rater, batch = gt_batch(6, id = "call_id"))
batch_fit <- gt_fit(batched, "quality", batch_design, control = control)
stopifnot(isTRUE(batch_fit$numerically_accepted))
within <- c("1" = 1, "2" = -1)
between <- c("1" = 1, "7" = -1)
corpus <- setNames(rep(1 / 24, 24), as.character(1:24))
batch_projection <- gt_batch_reliability(batch_fit,
  contrasts = list(within = within, between = between), aggregate = list(corpus = corpus))
subset(as.data.frame(batch_projection), target_kind != "absolute_item")
subset(batch_projection$source_contributions,
       source == "Call" & target_kind != "absolute_item")
```

A shared call shift cancels from the same-batch difference, remains in the
between-batch difference, and contributes to each absolute item score. A corpus
mean has another error target: call shifts average over calls, while common
rater shifts remain. Weighted aggregates can depend on how weights are spread
across groups. None of these effects can be represented by one universal batch
penalty.

```{r batch-design-study}
batch_study <- gt_batch_dstudy(batch_fit,
  expand.grid(rater = c(2, 4), run = c(1, 3)),
  contrasts = list(within = within, between = between), aggregate = list(corpus = corpus))
subset(as.data.frame(batch_study), target_kind == "contrast")
```

These are point projections conditional on the fitted grouping. Facet counts
change under exchangeability assumptions; item count, grouping, batch size and
fitted source covariances do not. All instrumentation facets remain random.
Aggregate error is annotation error conditional on the selected items, not
uncertainty from sampling a corpus. Contrast and aggregate ratios have explicit
names; they are not universal item-ranking G coefficients. Sampling intervals
are not provided.

Ordinary `gt_reliability()`, `gt_dstudy()` and the monetary `gt_plan()` refuse
modelled `Call` fits. The separate batch functions make their target explicit.
A fit keeps the compact item-to-batch labels even with data retention disabled;
projection objects contain those labels and target weights. Portable
`gt_report(fit)` and HTML output exclude them.

## Examine pilot precision for one final protocol

More pilot information may improve coefficient estimation without changing the
protocol eventually used for every item. The example varies pilot item and
rater counts, but every refit estimates the same four-rater, three-run target.

```{r pilot-precision}
precision <- gt_pilot_plan(fit,
  grid = data.frame(n_items = c(16, 24), rater = c(3, 4)),
  target_counts = c(rater = 4, run = 3),
  nsim = 2, seed = 2026, coefficient = "Phi", width_target = 0.25)
precision$summary
precision$replicates[, c("design_id", "replicate_id", "status", "accepted",
                         "interval_available", "interval_conditional", "width")]
```

Two simulations per candidate only exercise the software. They cannot support
a pilot-size recommendation. Set a predeclared simulation budget and inspect
Monte Carlo uncertainty before using the results for planning. Every attempted
replicate remains in the ledger. Precision success requires an available
interval meeting the width target, so errors, refusals and unavailable intervals
are non-successes. Mean widths condition on an available interval and show that
denominator. Conditional boundary intervals retain their qualification.

The simulator draws new items and new levels of every random facet from a fitted
or explicitly supplied Gaussian parameter scenario. It does not calibrate
coverage, integrate parameter uncertainty automatically, or provide bootstrap
confidence intervals. The initial simulation scope excludes batch declarations
and discrete outcomes. A model-based precision exercise does not establish
study adequacy when its scenario or facet universe is inappropriate.
